When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents

### 2024-10-10 article.Rmd
format changes JPCS

---
title: "When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents"

author:
  "Ismail Guennouni^1^*, Georgia Koppe^2^, Christoph Korn^3^"

abstract: |
  Can exposure to forgiving behavior improve cooperative outcomes, and does Rejection Sensitivity (RS) moderate this effect?  This randomized controlled experiment explores these questions using a novel approach with Hidden Markov Model (HMM)-based artificial agents in repeated trust games (RTGs). Participants (N = 206), pre-screened for high or low RS, interacted with either consistently behaving or more forgiving co-players, which were in reality HMM agents. Unexpectedly, exposure to forgiving agents led to reduced cooperation in subsequent interactions. Control group participants maintained consistent behavior. RS levels influenced perceptions of the agent's cooperativeness but did not significantly affect behavioral outcomes, revealing a perception-behavior dissociation. These findings challenge assumptions about fostering cooperation through simple exposure to positive interactions and highlight the importance of accounting for negative contrast effects when designing such interventions. They also demonstrate the potential of HMM-based artificial agents for studying interactive social dynamics, offering a methodological advancement for future research. Our results suggest the need for nuanced interventions that gradually shape expectations and behaviors to foster cooperative outcomes, particularly for individuals with high rejection sensitivity.

bibliography: "bib/exposure.bib"
biblio-style: spphys
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
header-includes:
  - \usepackage{lscape}
---
\small
^1^ *Central Institute of Mental Health, Medical Faculty Mannheim/Heidelberg University, Germany*

^2^ *Faculty of Mathematics and Computer Science, Heidelberg University, Germany*

^3^ *Department of General Psychiatry, Section Social Neuroscience, Heidelberg University, Germany*

^\*^ *Corresponding author. Address: Central institute of Mental Health, J5, Mannheim, Germany. Email: ismail.guennouni@zi-mannheim.de* <br>

@@

**Keywords** : Social Trust; Forgiveness Intervention; Rejection Sensitivity; Interpersonal Dynamics; Hidden Markov Models

@@

**Highlights:**

- Exposure to forgiving artificial agents unexpectedly led to reduced cooperation in subsequent interactions, challenging assumptions about fostering trust through positive experiences

- Rejection sensitivity influenced perceptions of artificial agents' cooperativeness but not behavioral outcomes, revealing a dissociation between perception and behavior

- Novel Hidden Markov Model-based artificial agents provide adaptive, human-like behavior in trust games, allowing for precise control of "cooperative" and "forgiving" strategies while maintaining experimental control of co-player strategies

@@

One explanation for such social challenges lies in early caregiver experiences. Attachment theory [@bowlby_attachment_1978] suggests that the quality of these relationships shapes our capacity for secure attachments and trust. Individuals with higher levels of insecure attachment may recall negative trust-related experiences more easily, report fewer positive trust experiences, and use less constructive coping strategies when trust is broken [@mikulincer_attachment_1998]. These insecure attachment patterns are often associated with heightened rejection sensitivity (RS), a tendency to anxiously expect, readily perceive, and intensely react to rejection [@downey_implications_1996; @downey_early_1997]. RS has been linked to the development of various mental health conditions, including depression, anxiety, personality disorders, and self-harm [@gao_associations_2017]. A recent meta-analysis revealed prosocial behavior and interpersonal trust as two key processes of interpersonal functioning that are markedly impaired in PDs and which are likely to contribute to interpersonal dysfunction in this population [@hepp_prosociality_2022]. The interaction of rejection sensitivity and trust-based constructs has been explored, particularly in Borderline Personality Disorder (BPD). @miano_rejection_2013 and @richetin_emotional_2018 found that rejection sensitivity mediated the relationship between BPD features and lower trust appraisal. @abramov_influence_2022-1 found that higher baseline feelings of rejection in individuals with BPD predict slower trust formation and less pronounced declines in trust following trust violations during the trust game. However, the interaction between trustworthiness and rejection sensitivity hasn't been studied as extensively in other personality disorders or in non-clinical populations, leaving a gap in our understanding of how these constructs might interplay. Could an ingrained sensitivity to rejection contribute to reduced pro-social behavior such as a failure to show trustworthiness?

If indeed RS contributes to interpersonal dysfunction through failure to show trustworthiness, could exposing those who exhibit it to cooperative and forgiving interaction partners lead to improved interpersonal outcomes? Research in the fields of behavioral economics and psychology has explored how positive social interactions influence cooperation and pro-social behavior. The use of the repeated trust game (RTG), a well-established experimental approach, has allowed for the analysis of the development of trust through ongoing interactions [@joyce_trust_1995]. In this paradigm, cycles of mutual trust, where each party's trust is reciprocated with trustworthiness, have the effect of enhancing cooperative behaviors even among individuals who are initially inclined to be distrustful [@king-casas_getting_2005]. @fowler_cooperative_2010 studied behavior in social networks interacting in a public goods game and found that cooperative behavior tends to cluster, suggesting that exposure to cooperative peers can lead to more cooperative behavior. Similarly, research on social learning theory [@bandura_social_1977] has long demonstrated that individuals learn and model the behavior of those around them, indicating that if someone is consistently exposed to cooperative and positive individuals, they're likely to emulate this behavior. These insights highlight that engaging with compassionate and forgiving others can be an effective method for mitigating deeply ingrained mistrust and potentially fostering trustworthy behavior.

In this study, we use a randomized controlled online experiment to test whether exposing participants with varying RS levels to forgiving and more cooperative co-players results in more trustworthy behavior and a repair of potential breakdowns in RTG cooperation. To simulate realistic social interaction while maintaining a high degree of experimental control, we take a novel paradigmatic approach: We use generative models of how humans play the RTG to design an agent that plays the role of the investor, based on Hidden Markov Models (HMMs) fitted to real players' data. A key aspect of these agents is that their actions depend on a latent "trust state" which reacts dynamically to the trustees' returns, simulating real-life trust-building scenarios. An advantage of having such a generative model of behavior is the possibility of controlling different aspects of the agent's strategy such as its general policy, the propensity to cooperate actively, or the propensity to retaliate after breakdowns of cooperation.

We pre-screened participants for high or low rejection sensitivity (RS) using a validated questionnaire, then assigned them exclusively to the trustee role in a series of trust games. After playing a 15-round RTG with a human-like HMM investor, they were randomly assigned to either a Control or Manipulation condition. In the Manipulation condition, participants were exposed over three RTGs to HMM investors designed with a limited propensity for retaliation, potentially mitigating ingrained mistrust. In the Control condition, participants played three RTGs against the same human-like HMM. After this exposure phase, all participants played another 15-round RTG with a human-like HMM investor. We expect those exhibiting high RS to be more responsive to the pre-programmed change in the agent's cooperativeness and retaliation propensity. We also hypothesize that those in the Manipulation condition would behave more cooperatively and have a lower propensity to retaliate to the co-player's defection after the exposure phase, with those in the higher RS group showing a bigger effect.

@@

To have participants with large differences in RS, a total of 1195 participants were pre-screened on the Prolific Academic platform (prolific.co) using the Rejection Sensitivity Questionnaire (RSQ) to finally select two similarly sized groups: One with high RS (RSQ score > 15) and the other with low RS (RSQ score < 10) totalling 206 participants (56% female). These were then invited through prolific to take part in the main experiment study. The required sample size was determined using an a priori power analysis to have an 80% probability to detect a small effect size with a 5% type I error rate in a repeated measures ANOVA with between factors. The mean age of participants was 34.6 years, with an 11.9 years standard deviation. Participants were paid a fixed fee of £6 plus a bonus payment dependent on their performance that averaged £0.5.

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS : High or Low) by 2 (Game: Trust-Game Pre Manipulation, Trust-Game Post Manipulation) design, with repeated measures on the third factor (Figure \@ref(fig:HMMPanels).A). Participants within each pre-screened group were randomly assigned to one of the two levels of the Condition factor. The games were designed and implemented online using Empirica v1 [@almaatouq_empirica_2021]. The planned experiment received approval from the University of Heidelberg's Medical Faculty ethics commission (ID:S-708/2023) and the experiment was performed in accordance with the ethics board guidelines and regulations. All participants provided informed consent prior to their participation.

@@

### Repeated Trust Game

@@

In all phases of the RTG other than the ‘Exposure phase’ (Figure \ref{fig:HMMPanels}.A), participants interacted with this human-like HMM. In the ‘Manipulation’ Condition of the exposure phase, however, the parameters of this HMM were adjusted to design a ‘forgiving’ and ultimately more cooperative agent. To achieve this, we changed the state transition probabilities of the HMM such that it becomes impossible for it to remain in a low trust state, effectively setting the transition probability for remaining in a “low-trust” state to 0. The resulting transition function is shown in Figure \ref{fig:HMMPanels}.D. The policies conditional on the latent states and the transition function in the other latent states remain unchanged.

\begin{landscape}
```{r HMMPanels, fig.cap= "\\small{A: A timeline of the experiment for both conditions. The RTG is played in dyads, with participants always assigned the role of the trustee and the HMM agent that of the investor. The investor is endowed with 20 units at the start of each round. They need to decide how much of that endowment they want to invest with the trustee. The investment is then multiplied by a factor of 3 and sent to the trustee who needs to decide how much of the multiplied investment they want to send back to the investor. The difference between conditions is the type of agents participants are exposed to in the exposure phase. Panels B - D: We construct the artificial investor agent by fitting a three-state hidden Markov model to data of human investors engaged in the 10 round RTG against human trustees. From the fitted HMM, we get the distribution of investments by the human-like agent, conditional on its latent state as shown in Panel B. The fitted HMM also yields the transition probability of the agent to a state on trial t+1 as a function of the net return (difference between the investment sent and the amount received in return) on trial t as shown in Panel C. Each plot in Panel C represents a different starting latent state on trial t, and each line represents the probability of transitioning to a particular state in trial t+1. Panel D shows the transition probabilities of the forgiving HMM agent in the low trust state. Unlike the human-like HMM, the forgiving HMM always transitions out of the low-trust state, and is more likely to end up in a high-trust state, as a proxy for coaxing behavior. Transitions in the medium and high trust states were identical for both agents.}",include=T, out.width='1\\linewidth', out.height='1\\textheight', fig.align='center'}

@@

# Display the combined image
plot(combined_image)
```
\end{landscape}

At the beginning of each game in all three phases, participants were told they would face a new player and had to wait to be paired with an available co-player. This simulated the waiting time in real social interaction tasks. After completing each RTG in each phase, participants rated how cooperative and forgiving they perceived the co-playerr to be, and whether they would like to play with them again (all on a scale from 1 to 10 with 10 being the most positive rating). After completing the three game phases, participants then completed the Levels of Personality Functioning Scale Brief-Form (LPFS-BF) questionnaire [@weekers_level_2019]. This is a self-report measure designed to assess core elements of personality functioning as defined in the Alternative Model for Personality Disorders in the DSM-5 [@american_psychiatric_association_diagnostic_2013], and provides a dimensional assessment of personality functioning, which complements the categorical approach of RS. Finally, participants were asked whether they thought the other players were human or computer agents, then debriefed and thanked for their participation.

@@

To explore whether participants behaved differently in the RTG after the manipulation compared to the Control group, we model the percentage return (percentage of tripled investment returned to investor) using a linear mixed-effects model with Phase (RTG game pre vs. post-manipulation), Condition (manipulation vs. Control), Investment, and RS (High vs Low RS group) as well as their interactions as fixed effects, and player-wise random intercepts and slopes for Phase. The full specification of the statistical model can be found in the supplement. To test whether participants behaved differently between conditions in the Exposure phase, we fit a linear mixed effects model to participants returns only in the exposure phase, with Condition (manipulation vs. Control), Investment, and RS (High vs Low RS group) as well as their interactions as fixed effects, and player-wise random intercepts. Finally, to test whether the HMM agent's behavior differed between Phases, Conditions and RS groups, we estimate a linear mixed-effects model of investments sent by the computerised HMM agent with Condition, Phase and RS and their interaction as fixed effects, and a similar random effects structure to the returns model.

The model was estimated using the `afex` package [@singmann_afex_2022] in R. More complex models with additional random effects could not be estimated reliably, and as such the estimated model can be considered to include the optimal random effects structure [@matuschek_balancing_2017]. A similar process was used to establish the random effects structures of linear mixed-effects models used to analyse the HMM agent investments as well as the participants' ratings of the co-players. There is no agreed upon way to calculate effect sizes for mixed effects models. Instead, we will report on testing differences in marginal means. For the $F$-tests, we used the Kenward-Roger approximation to the degrees of freedom, as implemented in the R package "afex". We Z-transform the Investment variable (subtract the overall investment mean and divide by overall standard deviation) as centering is beneficial to interpreting the main effects more easily in the presence of interactions.

@@

@@

<!-- EFFECT SIZES -->
```{r}
# Load the necessary packages

@@

On average, investments and returns, as shown in Figure \ref{fig:gamesPlot}, fell within the documented range of 40-60% of the endowment for investments and 35-50% of the total yield for returns, as reported in previous studies [@charness_investment_2008; @fiedler_social_2011].

@@

Mixed-effects analysis on the percentage returns shows a significant main effect of Phase (Pre vs. Post RTG game), `r papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f`, with higher percentage returns in the Pre RTG compared to the Post. Importantly, we also find an interaction between Condition and Phase (RTG pre- vs. post-manipulation), `r papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_condition_f`. As shown in Figure \ref{fig:boxPlots}, post-hoc tests confirm that, contrary to our expectations, there was a decrease in the percentage returned only in the Manipulation condition, pre - post,
`r papaja::apa_print(prepost_bycond)$full_result$Manipulation_Pre_post`, but no change in the Control condition.  We find no interaction between Phase, Condition and RS, suggesting that there was no difference between RS groups for this interaction.

There was also a significant main effect of Investment, 
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$inv_scaled)`,
such that higher investments were associated with higher percentage returns indicating positive reciprocity. An Investment by Condition interaction, 
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$condition_f_inv_scaled)`, reflected that returns were more affected by investments in the Control condition. We also find a three way interaction between Phase, Investment and RS, `r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_inv_scaled_high_RS)`, showing that the differentiated effect of the investment on the proportion returned by RS group is itself moderated by the Phase (pre- vs post manipulation). Finally, we find a four-way interaction between Condition, Phase, Investment and RS `r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_condition_f_inv_scaled_high_RS)`. For completeness, these effects are discussed in more detail in the supplement.

@@

@@

<!-- Focusing on post defection only:  -->
### Post Defection Trials

@@

```
```{r}
# Filter for expo games only
filtered_expo <- d_long %>%
  filter(gameNum.f %in% c("expo1", "expo2", "expo3"))

>>>>

---
title: "When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents"
author:
  "Ismail Guennouni^1,2,3,4^*, Georgia Koppe^1,2,3^$^\\dagger$, Christoph Korn^4^$^\\dagger$"
abstract: |
bibliography: "bib/exposure.bib"
csl: apa.csl
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
header-includes:
  - \usepackage{lscape}
editor_options: 
  markdown: 
    wrap: 72
---

\small

^1^ *Department of Psychiatry and Psychotherapy, Central Institute of
Mental Health, Medical Faculty Mannheim, Heidelberg University,
Mannheim, Germany*

^2^ *Interdisciplinary Center for Scientific Computing, Faculty of
Mathematics and Computer Science, Heidelberg University, Heidelberg,
Germany*

^3^ *Hector Institute for AI in Psychiatry, Central Institute of Mental
Health, Medical Faculty Mannheim, Heidelberg University, Mannheim
Germany*

^4^ *Department of General Psychiatry, Section Social Neuroscience,
Heidelberg University, Germany*

$^\dagger$ *Joint last author*

^\*^ *Corresponding author. Address: Central institute of Mental Health,
J5, Mannheim, Germany. Email:
[ismail.guennouni\@zi-mannheim.de](mailto:ismail.guennouni@zi-mannheim.de){.email}*

**Abstract:**

Can exposure to forgiving behavior improve cooperative outcomes, and
does Rejection Sensitivity (RS) moderate this effect? This randomized
controlled experiment explores these questions using a novel approach
with Hidden Markov Model (HMM)-based artificial agents in repeated trust
games (RTGs). Participants (N = 206), pre-screened for high or low RS,
interacted with either consistently behaving or more forgiving
co-players, which were in reality HMM agents. Unexpectedly, exposure to
forgiving agents led to reduced cooperation in subsequent interactions.
Control group participants maintained consistent behavior. RS levels
influenced perceptions of the agent's cooperativeness but did not
significantly affect behavioral outcomes, revealing a
perception-behavior dissociation. These findings challenge assumptions
about fostering cooperation through simple exposure to positive
interactions and highlight the importance of accounting for negative
contrast effects when designing such interventions. They also
demonstrate the potential of HMM-based artificial agents for studying
interactive social dynamics, offering a methodological advancement for
future research. Our results suggest the need for nuanced interventions
that gradually shape expectations and behaviors to foster cooperative
outcomes, particularly addressing the discrepancy between perceptions
and behavior in individuals with high rejection sensitivity.

@@

**Keywords** : Interpersonal functioning; Rejection Sensitivity;
Forgiveness Intervention; Trust-based Cooperation; Hidden Markov Models

@@

**General Scientific Summary:**

This study reveals that exposing individuals to forgiving behavior in
economic games can unexpectedly decrease their subsequent cooperation.
Interestingly, while people with high rejection sensitivity more
accurately perceived their partners' cooperativeness, this didn't
translate into more cooperative behavior. These findings challenge
assumptions about fostering cooperation and highlight the need for more
nuanced interventions to promote reciprocity, especially for individuals
prone to expecting rejection.

@@

One explanation for such social challenges lies in early caregiver
experiences. Attachment theory [@bowlby_attachment_1978] suggests that
the quality of these relationships shapes our capacity for secure
attachments and trust. Individuals with higher levels of insecure
attachment may recall negative trust-related experiences more easily,
report fewer positive trust experiences, and use less constructive
coping strategies when trust is broken [@mikulincer_attachment_1998].
These insecure attachment patterns are often associated with heightened
rejection sensitivity (RS), a tendency to anxiously expect, readily
perceive, and intensely react to rejection [@downey_implications_1996;
@downey_early_1997]. RS has been linked to the development of various
mental health conditions, including depression, anxiety, personality
disorders, and self-harm [@gao_associations_2017]. Individuals high in
RS show attentional biases towards social threat cues, which may
contribute to difficulties in social interactions
[@berenson_rejection_2009]. A recent meta-analysis revealed prosocial
behavior and interpersonal trust as two key processes of interpersonal
functioning that are markedly impaired in PDs and which are likely to
contribute to interpersonal dysfunction in this population
[@hepp_prosociality_2022]. The interaction of RS and trust-based
constructs has been explored, particularly in Borderline Personality
Disorder (BPD). @miano_rejection_2013 and @richetin_emotional_2018 found
that RS mediated the relationship between BPD features and lower trust
appraisal. @abramov_influence_2022-1 found that higher baseline feelings
of rejection in individuals with BPD predict slower trust formation and
less pronounced declines in trust following trust violations during the
trust game. However, the interaction between *reciprocity* and RS hasn't
been studied as extensively, leaving a gap in our understanding of how
these constructs might interplay. Could an ingrained sensitivity to
rejection contribute to reduced pro-social behavior such as a failure to
show reciprocity as a result of perceived rejection by the interaction
partner?

Given that RS may be a manifestation of maladaptive attachment styles,
it's crucial to explore whether exposure to consistently forgiving
interaction partners could help reshape these interpersonal expectations
and behaviors. If RS indeed influences reciprocity in repeated
interactions, could such exposure serve as a corrective experience,
potentially improving interpersonal outcomes? This approach aligns with
attachment theory, which suggests that new, positive relational
experiences can modify internal working models of relationships
[@bowlby_secure_1988]. Research in the fields of behavioral economics
and psychology has also explored how positive social interactions
influence cooperation and pro-social behavior. The use of the repeated
trust game (RTG), a well-established experimental approach, has allowed
for the analysis of the development of trust through ongoing
interactions [@joyce_trust_1995]. In this paradigm, cycles of mutual
trust, where each party's trust is reciprocated with trustworthiness,
have the effect of enhancing cooperative behaviors even among
individuals who are initially inclined to be distrustful
[@king-casas_getting_2005]. @fowler_cooperative_2010 studied behavior in
social networks interacting in a public goods game and found that
cooperative behavior tends to cluster, suggesting that exposure to
cooperative peers can lead to more cooperative behavior. Similarly,
research on social learning theory [@bandura_social_1977] has long
demonstrated that individuals learn and model the behavior of those
around them, indicating that if someone is consistently exposed to
cooperative and positive individuals, they're likely to emulate this
behavior. These insights highlight that engaging with compassionate and
forgiving others can be an effective method for mitigating deeply
ingrained mistrust and potentially fostering trustworthy behavior.

In this study, we use a randomized controlled online experiment to test
whether exposing participants with varying RS levels to forgiving and
more cooperative co-players results in more trustworthy behavior and a
repair of potential breakdowns in RTG cooperation. To simulate realistic
social interaction while maintaining a high degree of experimental
control, we take a novel paradigmatic approach: We use generative models
of how humans play the RTG to design an agent that plays the role of the
investor, based on Hidden Markov Models (HMMs) fitted to real players'
data. A key aspect of these agents is that their actions depend on a
latent "trust state" which reacts dynamically to the trustees' returns,
simulating real-life trust-building scenarios. An advantage of having
such a generative model of behavior is the possibility of controlling
different aspects of the agent's strategy such as its general policy,
the propensity to cooperate actively, or the propensity to trust again
after breakdowns of cooperation. To further mimic real-world
interactions and examine participants' responses to one-off breakdowns
of cooperation, we incorporate occasional pre-programmed low investments
by the agent.

We pre-screened participants for high or low RS using a validated
questionnaire, then assigned them exclusively to the trustee role in a
series of trust games. After playing a 15-round RTG with a human-like
HMM investor, they were randomly assigned to either a Control or
Manipulation condition. In the Manipulation condition, participants were
exposed over three RTGs to HMM investors designed with a limited
propensity for retaliation, potentially mitigating ingrained mistrust.
In the Control condition, participants played three RTGs against the
same human-like HMM. After this exposure phase, all participants played
another 15-round RTG with a human-like HMM investor, similar to the one
in the pre-exposure phase. We expect those exhibiting high RS to be more
responsive to the pre-programmed change in the agent's cooperativeness
and retaliation propensity. We also hypothesize that those in the
Manipulation condition would behave more cooperatively and have a lower
propensity to retaliate to the co-player's defection after the exposure
phase, with those in the higher RS group showing a bigger effect.

@@

To have participants with large differences in RS, a total of 1195
participants were pre-screened on the Prolific Academic platform
(prolific.co) using the Rejection Sensitivity Questionnaire (RSQ) to
finally select two similarly sized groups: One with high RS (RSQ score
\> 15, N=103) and the other with low RS (RSQ score \< 10, N=103)
totalling 206 participants (56% female). These were then invited through
prolific to take part in the main experiment. The required sample size
was determined using an *a priori* power analysis to have an 80%
probability to detect a small effect size (Cohen’s f = 0.10) for a
within-between interaction with a 5% type I error rate in a repeated
measures ANOVA. The sample size calculation assumed 2 groups, 2
measurement per group and was performed using the G\*Power software
[@faul_statistical_2009]. The mean age of participants was 34.6 years,
with an 11.9 years standard deviation. The majority of participants
identified ethnically as White ($80$%). The online cohort registered 30
unique countries of birth with the most frequent being the U.K ($33$%)
followed by Poland ($10$%) and Portugal ($10$%). Participants were paid
a fixed fee of £6 plus a bonus payment dependent on their performance
that averaged £0.5. Data was collected over multiple sessions between
December 2023 and February 2024.

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS :
High or Low) by 2 (Phase: Trust-Game Pre Manipulation, Trust-Game Post
Manipulation) design, with repeated measures on the Phase factor (Figure
\@ref(fig:HMMPanels).A). Participants within each pre-screened RS group
were randomly assigned to one of the two levels of the Condition factor,
resulting in 101 participants in the Manipulation condition and 105 in
the Control condition. The games were designed and implemented online
using Empirica v1 [@almaatouq_empirica_2021]. The planned experiment
received approval from the University of Heidelberg's Medical Faculty
ethics commission (ID:S-708/2023) and the experiment was performed in
accordance with the ethics board guidelines and regulations. All
participants provided informed consent prior to their participation.

@@

### Repeated Trust Game and HMM Investor

@@

In all phases of the RTG other than the ‘Exposure phase’ (Figure
\@ref(fig:HMMPanels).A), participants interacted with this human-like
HMM. In the ‘Manipulation’ Condition of the exposure phase, however, the
parameters of this HMM were adjusted to design a ‘forgiving’ and
ultimately more cooperative agent. To achieve this, we changed the state
transition probabilities of the HMM such that it becomes impossible for
it to remain in a low trust state, effectively setting the transition
probability for remaining in a “low-trust” state to 0. The resulting
transition function is shown in Figure \@ref(fig:HMMPanels).D. The
policies conditional on the latent states and the transition function in
the other latent states remain unchanged.

At the beginning of each game in all three phases, participants were
told they would face a new player and had to wait to be paired with an
available co-player. This simulated the waiting time in real social
interaction tasks. After completing each RTG in each phase, participants
rated how cooperative and forgiving they perceived the co-player to be,
and whether they would like to play with them again (all on a scale from
1 to 10 with 10 being the most positive rating). After completing the
three game phases, participants then completed the Levels of Personality
Functioning Scale Brief-Form (LPFS-BF) questionnaire
[@weekers_level_2019]. This is a self-report measure designed to assess
core elements of personality functioning as defined in the Alternative
Model for Personality Disorders in the DSM-5
[@american_psychiatric_association_diagnostic_2013], and provides a
dimensional assessment of personality functioning, which complements the
categorical approach of RS. Finally, participants were asked whether
they thought the other players were human or computer agents, to probe
how well the agent can mimic human behavior, then debriefed and thanked
for their participation.

````{=tex}
\begin{landscape}
```{r HMMPanels, fig.cap= "\\small{A: A timeline of the experiment for both conditions. The RTG is played in dyads, with participants always assigned the role of the trustee and the HMM agent that of the investor. The investor is endowed with 20 units at the start of each round. They need to decide how much of that endowment they want to invest with the trustee. The investment is then multiplied by a factor of 3 and sent to the trustee who needs to decide how much of the multiplied investment they want to send back to the investor. The difference between conditions is the type of agents participants are exposed to in the exposure phase. Panels B - D: We construct the artificial investor agent by fitting a three-state hidden Markov model to data of human investors engaged in the 10 round RTG. From the fitted HMM, we get the distribution of investments by the human-like agent, conditional on its latent state as shown in Panel B. The fitted HMM also yields the transition probability of the agent to a state on trial t+1 as a function of the net return (difference between the investment sent and the amount received in return) on trial t as shown in Panel C. Each plot in Panel C represents a different starting latent state on trial t, and each line represents the probability of transitioning to a particular state in trial t+1. Panel D shows the transition probabilities of the forgiving HMM agent in the low trust state. Unlike the human-like HMM, the forgiving HMM always transitions out of the low-trust state, and is more likely to end up in a high-trust state, as a proxy for coaxing behavior. Transitions in the medium and high trust states were identical for both agents.}",include=T, out.width='1\\linewidth', out.height='1\\textheight', fig.align='center'}

@@

# Display the combined image
plot(combined_image)
```
\end{landscape}
````

@@

To test whether participants behaved differently in the RTG after the
manipulation compared to the Control group, we model the percentage
return (percentage of tripled investment returned to investor) using a
linear mixed-effects model with Phase (RTG game pre vs.
post-manipulation), Condition (manipulation vs. Control), Investment,
and RS (High vs Low RS group) as well as their interactions as fixed
effects, and player-wise random intercepts and slopes for Phase. The
full specification of the statistical model can be found in the
supplement. To test whether participants behaved differently between
conditions in the Exposure phase, we fit a linear mixed effects model to
participants returns only in the exposure phase, with Condition
(manipulation vs. Control), Investment, and RS (High vs Low RS group) as
well as their interactions as fixed effects, and player-wise random
intercepts. Finally, to test whether the HMM agent's behavior differed
between Phases, Conditions and RS groups, we estimate a linear
mixed-effects model of investments sent by the computerised HMM agent
with Condition, Phase and RS and their interaction as fixed effects, and
a similar random effects structure to the returns model.

The model was estimated using the `afex` package [@singmann_afex_2022]
in R. More complex models with additional random effects could not be
estimated reliably, and as such the estimated model can be considered to
include the optimal random effects structure
[@matuschek_balancing_2017]. A similar process was used to establish the
random effects structures of linear mixed-effects models used to analyse
the HMM agent investments as well as the participants' ratings of the
co-players. There is no agreed upon way to calculate effect sizes for
mixed effects models. Instead, we will report on testing differences in
marginal means. For the $F$-tests, we used the Kenward-Roger
approximation to the degrees of freedom, as implemented in the R package
"afex". We Z-transform the Investment variable (subtract the overall
investment mean and divide by overall standard deviation) as centering
is beneficial to interpreting the main effects more easily in the
presence of interactions. To probe significant interactions, we
conducted planned contrasts using the `emmeans` package in R. Given that
we were testing multiple pre-planned comparisons, we applied the "Sidak"
correction to control for familywise error rate while maintaining
reasonable statistical power. This approach allowed us to investigate
specific hypotheses about the differential effects of our manipulation
across phases and RS groups, while protecting against inflated Type I
error rates.

@@

```{r}
# Number of participants in manipulation condition
temp_manip <- avg_ret_df %>% filter(condition.f =="Manipulation")

length(unique(temp_manip$playerId))

# Number of participants with high RS

temp_RS <- avg_ret_df %>% filter(high_RS =="high RS")
length(unique(temp_RS$playerId))

@@

<!-- EFFECT SIZES -->

```{r}
# Load the necessary packages

@@

On average, investments and returns, as shown in Figure
\@ref(fig:gamesPlot), fell within the documented range of 40-60% of the
endowment for investments and 35-50% of the total yield for returns, as
reported in previous studies [@charness_investment_2008;
@fiedler_social_2011].

@@

Mixed-effects analysis on the percentage returns shows a significant
main effect of Phase (Pre vs. Post RTG game),
`r papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f`, with
higher percentage returns in the Pre RTG compared to the Post.
Importantly, we also find an interaction between Condition and Phase
(RTG pre- vs. post-manipulation),
`r papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_condition_f`.
As shown in Figure \@ref(fig:boxPlots), post-hoc tests confirm that,
contrary to our expectations, there was a decrease in the percentage
returned only in the Manipulation condition, pre - post,
`r papaja::apa_print(prepost_bycond)$full_result$Manipulation_Pre_post`,
but no change in the Control condition. We find no interaction between
Phase, Condition and RS, suggesting that there was no difference between
RS groups for this interaction.

There was also a significant main effect of Investment,
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$inv_scaled)`,
such that higher investments were associated with higher percentage
returns indicating positive reciprocity. An Investment by Condition
interaction,
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$condition_f_inv_scaled)`,
reflected that returns were more affected by investments in the Control
condition. We also find a three way interaction between Phase,
Investment and RS,
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_inv_scaled_high_RS)`,
showing that the differentiated effect of the investment on the
proportion returned by RS group is itself moderated by the Phase (pre-
vs post manipulation). Finally, we find a four-way interaction between
Condition, Phase, Investment and RS
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2", papaja::apa_print(mod_returns_pct_RS)$full_result$phase_f_condition_f_inv_scaled_high_RS)`.

@@

```{r}
library(emmeans)

# First, get the estimated marginal means
emm <- emmeans(mod_returns_pct_RS, ~ inv_scaled | phase.f * condition.f * high_RS)

# Then, get the slopes (effects of investment) for each combination
slopes <- emtrends(mod_returns_pct_RS, ~ phase.f * condition.f * high_RS, var = "inv_scaled")
slopes

# Define the contrasts
corrected_contrasts <- list(
  "Diff in inv effect (Post - Pre, Manip, Low RS)" = c(0, 0, -1, 1, 0, 0, 0, 0),
  "Diff in inv effect (Post - Pre, Manip, High RS)" = c(0, 0, 0, 0, 0, 0, -1, 1),
  "Diff in (Post - Pre) between High and Low RS" = c(0, 0, 1, -1, 0, 0, -1, 1)
)

# Test the contrasts with multiple comparison correction
results_4w <- contrast(slopes, method = corrected_contrasts, adjust = "sidak")

# View the results
summary(results_4w)

To examine this four-way interaction, we conducted a contrast analysis
of how the effect of investment on returns changed from pre- to
post-manipulation for different RS groups in both conditions. Starting
with the Manipulation condition, for participants with low RS, the
effect of investment on returns increased significantly from pre- to
post-manipulation,
`r papaja::apa_print(results_4w)$full_result$DiffininveffectPost_PreManipLowRS`.
This suggests that after the manipulation, low RS participants became
more responsive to their co-player's investments, returning
proportionally more as investments increased. In contrast, for
participants with high RS, the effect of investment on returns decreased
significantly from pre- to post-manipulation,
`r papaja::apa_print(results_4w)$full_result$DiffininveffectPost_PreManipHighRS`.
This indicates that high RS participants became less responsive to their
co-player's investments after the manipulation, with smaller increases
in returns as investments increased. The difference in these pre-post
changes between high and low RS groups was significant,
`r papaja::apa_print(results_4w)$full_result$DiffinPost_PrebetweenHighandLowRS`.
This result suggests that the manipulation had significantly different
effects on how low and high RS participants responded to their
co-player's investments.

In the Control condition, we observed no significant changes in how
participants responded to their co-player's investments between the pre
and post phases, regardless of their RS level.

```{r}
# Define the contrasts for both conditions
contrasts_ctrl <- list(
  # Control condition contrasts
  "Diff in inv effect (Post - Pre, Control, Low RS)" = c(-1, 1, 0, 0, 0, 0, 0, 0),
  "Diff in inv effect (Post - Pre, Control, High RS)" = c(0, 0, 0, 0, -1, 1, 0, 0),
  "Diff in (Post - Pre) between High and Low RS in Control" = c(1, -1, 0, 0, -1, 1, 0, 0)
)

# Test the contrasts with multiple comparison correction
results_ctrl <- contrast(slopes, method = contrasts_ctrl, adjust = "sidak")

# View the results
print(summary(results_ctrl))
```

@@

<!-- Focusing on post defection only:  -->

### Post Defection Trials

@@

```{r}
# Filter for expo games only
filtered_expo <- d_long %>%
  filter(gameNum.f %in% c("expo1", "expo2", "expo3"))
